Microplastics in the Urban Environment

### 2021-10-26 code/analysis_code/exploration.Rmd
updated pop data, wrf data, land cover FIPS

>>>>

#Working on gathering WRF distance
```{r}
#set individual WRF locations and save as values
cedarcreeklat = 33.87351 
cedarcreeklong = -83.33639
northoconeelat = 33.93402
northoconeelong = -83.36018
middleoconeelat = 33.90862
middleoconeelong = -83.39214

mpdata <- mpdata %>% mutate(ccwrf = geodist_vec(mpdata$long, mpdata$lat, cedarcreeklong, cedarcreeklat))
mpdata <- mpdata %>% mutate(nowrf = geodist_vec(mpdata$long, mpdata$lat, northoconeelong, northoconeelat))
mpdata <- mpdata %>% mutate(miwrf = geodist_vec(mpdata$long, mpdata$lat, middleoconeelong, middleoconeelat))

#Since I only care about the proximity to the closest WRF, my next task is to write code that creates a new variable returning only the lowest value out of ccwrf, nowrf, and miwrf for each sample site.

#subset data into just distance from each WRF
wrf <- mpdata %>% select("ccwrf", "nowrf", "miwrf") 
#find minimum distance to any WRF out of all three
val <- do.call(pmin, c(wrf, na.rm = TRUE)) 
minwrf <- transform(wrf, min=val, minCol = names(wrf)[max.col(wrf == val, 'first')])

#minwrf contains the variable min, which tells us the distance from the sample site to the nearest WRF. This is all of the information that we need, so no we will re-join to the original DF
mpwrfdata <- left_join(mpdata, minwrf)

#remove variables we won't need anymore
mpwrfdata <- mpwrfdata %>% select(-"ccwrf", 
                                  -"nowrf",
                                  -"miwrf")
#rename columns
mpwrfdata <- mpwrfdata %>% rename("dist" = "min", 
                     "wrf" = "minCol")

#save as RDS
saveRDS(mpwrfdata, file = here("data", "processed_data", "mpwrfdata.rds"))
```

Update 10/25/21
I've updated the population data by accessing census tract level population rather than zip code level population. This should give us more localized population levels and might paint a better picture of the relationship between MP particles per liter and local population level.

#Adding WRF data to updated population data
```{r}
#load data
tractpopdata <- readRDS(here("data", "processed_data", "tractpopdata.rds"))

#add wrf data to tractpopdata
wrfjoin <- mpwrfdata %>% select("site", "date", "wrf", "dist") #selecting just the variables we want to add
mpwrfcombo <- left_join(tractpopdata, wrfjoin, by = c("site", "date")) #joining the two datasets

#save as RDS
saveRDS(mpwrfcombo, file = here("data", "processed_data", "mpwrfcombo.rds"))
```

Now, `mpwrfcombo` contains the sample info and the new WRF distance info. However, something to work on going forward is that a sample site may be near a WRf but not downstream of it. It might be more important to know the distance to the nearest upstream WRF. I'll have to go back and consider this as I continue my exploration/analysis.

Other than not including landcover data, `mpwrfcombo` is the most comprehensive dataset, including almost all of the data that I have compiled so far.

# Plotting population vs particles per liter w/ new data
```{r}
tractpopdata %>% ggplot(aes(x = population, y = particles_l)) + geom_point() 
```

With this plot, we still don't see an obvious relationship between particles per liter and population level, even with our updated population levels.

```{r}
tractpopdata %>% ggplot(aes(x = population, y = particles_l)) + geom_point() + scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')
```

### 2021-10-26 code/processing_code/censusland_code.Rmd
updated pop data, wrf data, land cover FIPS

#filter to clean data set
landcoverclean = landcover %>% select(-GISJOIN, -state_fip, -county_fip, -OID)

athensfips <- c("13219", "13059")
athensland <- filter(landcoverclean, fips %in% athensfips)

```

>>>>

#clean dataset by selecting only data for the state of Georgia
landcover <- landcover %>%
  filter(state_fip == "13")

#get rid of the 4th and the 8th zero in GISJOIN to make compatible with FIPS codes for samples + population count
landcover = landcover %>% mutate(FIPS = paste(state_fip, county_fip, substring(landcover$GISJOIN, 9, 14), sep = "")) %>% relocate(FIPS)

#filter to clean data set
landcoverclean = landcover %>% select(-GISJOIN, -state_fip, -county_fip, -OID, fips)

#Now we have landcover data in a form that is compatible with our sample site data (using small geography census info). Let's join datasets to create a dataset with sample site info and land cover info
tractpopdata <- readRDS(here("data", "processed_data", "tractpopdata.rds"))
athensland <- left_join(tractpopdata, landcoverclean, by = "FIPS")

#save as RDS
saveRDS(athensland, file = here("data", "processed_data", "athensland.rds"))
```

### 2021-10-26 code/processing_code/censusprocessing.Rmd
updated pop data, wrf data, land cover FIPS

>>>>

Update 10/25/21

I am realizing that we can get more localized population levels by using census tracts, which are smaller levels of census geography. Let's try to use FIPS codes (which include tract info) to obtain population data that might be more informative. 
# Converting coordinates to FIPS
```{r}
#create function
latlong2fips <- function(latitude, longitude) {
  url <- "https://geo.fcc.gov/api/census/block/find?format=json&latitude=%f&longitude=%f"
  url <- sprintf(url, latitude, longitude)
  json <- RCurl::getURL(url)
  json <- RJSONIO::fromJSON(json)
  as.character(json$Block['FIPS'])
}
#vectorize so function will take more than one argument
latlong2fips_vec <- Vectorize(latlong2fips, vectorize.args = "latitude", "longitude")

fips <- popdata %>% rowwise() %>% mutate(FIPS = latlong2fips_vec(lat, long))

saveRDS(fips, "fips.RDS")
```

# FIPS to census tract population
```{r}
library(tidycensus)
#tidycensus::census_api_key("c02ef0638e60f9d5fbb45de5219821e3a25cd7d2", install = TRUE)
get_decennial("block", state = Georgia, variables = population, year = 2019)
```

Here, we receive an error message suggesting to use NHGIS data since the most recent small census geography data is not available through `get_decennial()`.

# Using NHGIS data for more recent block-level population counts
```{r}
#Downloaded data from http://NHGIS.org
install.packages("ipumsr")
library(ipumsr)

#read in census tract level population data
tract <- read.csv(here("data","raw_data","nhgis0001_csv", "nhgis0001_ds244_20195_tract.csv"))

#clean up dataset to remove unneeded columns/variables
tract <- tract %>% filter(STATE == "Georgia") %>% select(1:3, 6:9, 12, 40, 43:46)

#Now, we have the FIPS codes for each sample site (down to the tract/block level), and we have population counts by FIPS code.

#First, we need to remove the last four digits of the FIPS code in the fips dataset so that it will correspond with the #FIPS code in the tract dataset. We'll also need to remove the G from the beginning of the tract FIPS codes.

#combine individual codes for state, county, tract to get long form FIPS code
a <- tract %>% mutate(FIPS = paste(STATEA, COUNTYA, sep = "")) %>%
  mutate(FIPS = paste(FIPS, TRACTA, sep = "0"))

#keep only the variables we need, FIPS code and population size (ALUBE001)
a <- a %>% select("FIPS", "ALUBE001")
#remove last four digits from fips codes associates with samples, so that they coordinate with population census tract FIPS codes
b <- fips %>% mutate(FIPS = (substr(FIPS,1,nchar(FIPS)-4)))
#combine datasets by FIPS code, so that population size is added to our main dataset based on corresponding FIPS code
blockpop <- left_join(b, a, by = "FIPS")

#We still have some missing values, but when I search for the FIPS code that are missing population levels, I am able to find them in the tract dataset. It looks like the FIPS code/GEOID are sometimes the same but sometimes differ by one digit (usually a zero inserted in the middle of the string somewhere). Let's adjust the GEOID variable to make it comparable to the FIPS codes we are still missing population data for, and then we'll join the datasets again to try to fill in missing values.

#let's try GEOID minus first 7 digits
tract$GEOID <- gsub("^.{0,7}", "", tract$GEOID) 
#selecting only the variables we want to add in our final combination dataset
tractpop <- tract %>% select("GEOID", "ALUBE001")
#joining based on FIPS code to fill in missing population data
c <- left_join(blockpop, tractpop, by = c("FIPS" = "GEOID"))

#This last attempt was successful and we now have a column that gives us the census tract level population count, with no missing values. We no longer need the first column, or the zip code and zip code population data that we looked at initially. Let's clean up this dataset for further exploration

#Remove first ALUBE001 column and zip, population columns
tractpopdata <- c %>% select(-"ALUBE001.x", -"zip", -"population")
tractpopdata <- rename(tractpopdata, "population" = "ALUBE001.y")

#save as RDS
saveRDS(tractpopdata, file = here("data", "processed_data", "tractpopdata.rds"))

### 2021-10-26 products/manuscript/Manuscript.Rmd
updated pop data, wrf data, land cover FIPS

---
title: "Microplastics in the Urban Environment"
subtitle: "Exploratory Study of Freshwater Microplastics in Athens, GA"
author: Amelia M. Foley
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
#bibliography: library.bib #I am having trouble with the bib file (error message: pandoc error 83 file not found). Removing this line for now so that Manuscript will knit. Will continue to troubleshoot and fix bib formatting. 
csl: ../apa.csl
---

>>>>

---
title: "Microplastics in the Urban Environment"
subtitle: "Exploratory Study of Freshwater Microplastics in Athens, GA"
author: Amelia M. Foley
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: here::here("products", "library.bib") #I am having trouble with the bib file (error message: pandoc error 83 file not found). Removing this line for now so that Manuscript will knit. Will continue to troubleshoot and fix bib formatting. 
csl: ../apa.csl
---
